---
title: Power calculation for COVID-19 and air pollution study
author: Sergio Olmos
date: August 2020
output:
  html_document:
    theme: cosmo
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r}
library(drake)
library(tidyverse)
```
Statistical power was assessed using the closed-form formula derived by Hsieh and Lavori (2000)[^1] for the Cox PH model as implemented in the `powerSurvEpi` R package[^2].

The calculations assume the following Cox PH model:

$$h(t | x_1, \boldsymbol{x}_2) = h_0(t) \exp(\beta_1 x_1 + \boldsymbol{\beta}_2 \boldsymbol{x}_2)$$

where the covariate $X_1$ represents PM2.5 exposure and $\boldsymbol{X}_2$ is a vector of other covariates.

Given the type I error rate $\alpha$ for a two-sided test, the power required to detect a hazard ratio as small as $\exp(\beta_1) = \theta$ is

$$power = \phi \left(-z_{1-\alpha/2} + \sqrt{n[\log(\theta)^2] \sigma^2 \psi (1 - \rho^2)}\right)$$

where $\sigma^2 = Var(X_1)$, $\psi$ is the proportion of events of interest, and $\rho$ is the multiple correlation coefficient of the following linear regression:

$$x_1 = b_0 + \boldsymbol{b}^T \boldsymbol{x}_2.$$

That is, $\rho^2 = R^2$, where $R^2$ is the proportion of variance explained by the regression of $X_1$ on the vector of covariates $\boldsymbol{X}_2$. The parameter $\rho^2$ can therefore be seen as the amount of additional information that PM2.5 provides once we know the value of other covariates. The smaller $\rho^2$ is, the more power our analysis will have for any given effect size.

In our power calculations we will consider the sample size to be fixed at 5 million and compute the power for a range of hazard ratios and $\rho^2$ values.

Incidence was obtained from Generalitat de Catalunya[^3] and Ministerio de Sanidad[^4].

## Exposure

The air pollution exposures used in the power calculations are derived from APACHE model estimates linked to over 3 million people in Cataluña from PADRIS.

For long-term exposure we used the variance in PM2.5 annual exposures for all 3 million individuals.

For short-term exposure, daily measurements from Vall d'Hebron station from 2014 to 2018 were used to estimate PM2.5 exposures variance. The estimated day-to-day variance is much larger than the spatial variance at any given time point as expected.


## Aim 1

Test whether long-term exposure to air pollution is associated with COVID-19 hospital admissions or mortality in the general population.

### Incidence

* Population size: ~5,000,000
* Number of COVID-19  cases that needed hospitalization: ~32,000
* Number of COVID-19 related deaths: ~13,000

### Exposure

For long-term exposure, the variance in PM2.5 exposure is taken to be the variance in average annual PM2.5 exposure for each individual, as estimated by APACHE:

* Variance: `r round(readd(pm25_var_aim1), 2)`

### Power

```{r}
plot_h_1 <- readd(hospi_plot_aim1) +
  ggtitle(
    "Power for detecting higher risk in hospitalizations\nfor long-term exposure",
    subtitle = "N = 5,000,000  Hospitalizations = 32,000\n"
  ) +
  ylab(bquote(rho^2))
plot_h_1
plot_d_1 <- readd(dead_plot_aim1) +
  ggtitle(
    "Power for detecting higher risk in mortality\nfor long-term exposure",
    subtitle = "N = 5,000,000 Deaths = 13,000\n"
  ) +
  ylab(bquote(rho^2))
plot_d_1

ggsave(
  plot = plot_h_1,
  filename = here("figs", "power-hosp-aim1.png"),
  width = 8, height = 6
)
ggsave(
  plot = plot_d_1,
  filename = here("figs", "power-dead-aim1.png"),
  width = 8, height = 6
)
```

## Aim 2

Test whether short-term exposure to air pollution is associated with COVID-19 hospital admissions or mortality in the general population.

### Incidence

* Number of COVID-19 diagnosed subjects: ~115,000
* Number of COVID-19 cases that needed hospitalizations: ~32,000
* Number of COVID-19 related deaths: ~13,000

### Exposure

For short-term exposure, PM2.5 exposure variance is taken to be the average variance in daily PM2.5 concentration measurements from Vall d'Hebron station among all recorded months:

* Variance: `r round(readd(pm25_var_aim2), 2)`

### Power

```{r}
plot_h_2 <- readd(hospi_plot_aim2) +
  ggtitle(
    "Power for detecting higher risk in hospitalizations\nfor short-term exposure",
    subtitle = "N = 115,000  Hospitalizations = 32,000\n"
  ) +
  ylab(bquote(rho^2))
plot_h_2
plot_d_2 <- readd(dead_plot_aim2) +
  ggtitle(
    "Power for detecting higher risk in mortality\nfor short-term exposure",
    subtitle = "N = 115,000  Deaths = 13,000\n"
  ) +
  ylab(bquote(rho^2))
plot_d_2

ggsave(
  plot = plot_h_2,
  filename = here("figs", "power-hosp-aim2.png"),
  width = 8, height = 6
)
ggsave(
  plot = plot_d_2,
  filename = here("figs", "power-dead-aim2.png"),
  width = 8, height = 6
)
```

## References

[^1]: Hsieh F.Y. and Lavori P.W. (2000). Sample-size calculation for the Cox proportional hazards regression model with nonbinary covariates. Controlled Clinical Trials. 21:552-560.

[^2]: powerSurvEpi: Power and Sample Size Calculation for Survival Analysis of Epidemiological Studies. (https://cran.r-project.org/web/packages/powerSurvEpi/index.html)

[^3]: https://dadescovid.cat/

[^4]: https://github.com/victorvicpal/COVID19_es/blob/master/data/pdf_data/Actualizacion_90_COVID-19.pdf
